Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-19 paper/paper.tex
Update SPDS and DPDS descriptions

\subsection{Static Provisioning Dynamic Scheduling\\
(SPDS) }
\label{sec:spds}
This is the simplest strategy based on static provisioning of resources and
on-line scheduling of workflow tasks. Given the budget $B$ and deadline $d$
together with the hourly price $p$ of a VM it is possible to compute the number
of VMs $N_{VM}$ to provision so that they consume the whole budget before the
deadline.

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil B / d / p \rceil
\end{equation}

Although simple, such provisioning plan has the advantage that it
minimizes the number of provisioning and deprovisioning requests.
Once the resources are provisioned, the tasks are scheduled to VMs using dynamic
priority-based Algorithm~\ref{alg:ds}.

\begin{algorithm}
\caption{Priority-based scheduling}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{schedule}{}
    \State $Q\gets set\ of\ released\ tasks$
    \State $P\gets empty\ priority\ queue$
    \While{$Q \neq \emptyset$}
    	\For{$task\ t\in Q$}
    		\State \Call{Insert}{$t,P$} 
    	\EndFor
    	\While{$FreeVMs \neq \emptyset \wedge P \neq \emptyset $}
    		\State $VM\gets$ \Call{SelectRandom}{$FreeVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,VM$}
    	\EndWhile
    	\State \Call{Wait}{$task\ finished$}
    	\State \Call{Update}{$Q$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

Initially all the
released tasks from all the workflows are submitted to the queue $Q$. Then, they
are inserted into priority queue $P$ based on the priority of workflows. If
there are free VMs available, the tasks from the priority queue are submitted to
the randomly choosen VMs. The process is repeated each time a task finishes.
Using the priority queue guarantees that tasks from lower priority workflows are
always submitted later than the higher priority ones and that the lower priority
tasks can occupy idle VMs. However, as there is no preemption, long-running
low-priority tasks may delay execution of high-priority tasks when such ones
become ready. Fig.~\ref{fig:spds-example} shows the example plan generated using
SPDS algorithm. It can be seen that tasks from lower priority workflows can fill
idle time of provisioned virtual machines.

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example of schedule generated using SPDS algorithm: tasks are labelled
 by colors depending on the workflow. }
\label{fig:spds-example}
\end{figure}

@@

Static provisioning plan used in SPDS algorithm allows no changes to resource
pool at runtime. However, such plan may not be optimal when resource demand
changes during execution of workflows. In policy-based provisioning
(or auto-scaling) systems the metrics such as resource
utilization or queue length are typicaly used to estimate curret resource
demand. Our DPDS algorithm is based on resource utilization: the provisioner
module periodically computes utilization as the percentage of idle VMs. The
provisioner algorithm detects which VMs are finishing their hourly billing cycle
during the next provisioning period and decides whether they should continue
running, should be terminated or new VMs should be provisioned: see
Algorithm~\ref{alg:prov}.

\begin{algorithm}
\caption{Dynamic provisioner}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization thershold; $u_l$: lower utilization
threshold.
\Procedure{provision}{}
    \State $V_C\gets set\ of\ VMs\ completing\ billing\ cycle$
    \State $n_T\gets\ 0$ \Comment{number of VMs to terminate} 
    \If{$ b - c < p * |V_C| \vee t > d $ }
    	\State $V_R\gets set\ of\ running\ VMs$
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$ 
    	\State $V_T\gets select\ n_T VMs\ to\ terminate\ from\ V_C$
    	\State \Call{Terminate}{$V_T$}
    \Else 
    	\If{$u>u_h$}
    		\State \Call{Start}{$new\ VM$}
    	\EndIf
    	\If{$u<u_l$}
    		\State $V_F\gets free\ (idle)\ VMs$
    		\State $n_T\gets \lceil|V_F|/2\rceil$ 
    		\State \Call{Terminate}{$V_T$}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The number of instances to terminate $n_T$ computed in line 6 is the number that
would overrun the budget if not terminated. On the other hand $n_T$ computed in
line 15 prevents terminating all the instances too quickly, but makes sure that
if there is only one instance left it should be terminated. However, the
termination process is not immediate: to make sure that VMs that have been
already paid for are not terminated prematurely the {\em Terminate()} procedure
terminates only these instances that are close to the end of their billing
cycle. The remaining time threshold is calculated as a sum of provisioner
interval and estimated termination delay. This guarantees that no VMs can
overrun the assigned budget. To avoid uncontrolled increase of number of
instances, which may happen in the case of highly parallel workflows,
provisioner checks also the maximum scaling factor from the initial number of
provisioned VMs and prohibits starting new ones when this threshold is reached.

@@

DPDS algorithm does not use any information about workflow structure: it deals
only with scheduling of individual released/ready tasks using their
workflow priorities. If there are no high priority tasks in the ready queue,
then lower priority tasks are submitted to idle resources. This may lead to the
deficiency in the following scenario: if a long-running task of low
priority is submitted to the idle machine and then several high
priority tasks are released to the ready queue, they become delayed by the
execution of the low priority task. This usually does not cause a problem when
the number of tasks and resources is high, but when the execution approaches the
budget and deadline constraint, we have observed that tasks of low-priority
workflows may prevent higher priority workflows from completion. This
observation leads to the improvemed algorithm which we call workflow-aware DPDS
(WA-DPDS).

>>>>

In this section we describe several algorithms we developed to execute ensembles of
workflows on the cloud under budget and deadline constraints.

\subsection{Static Provisioning Dynamic Scheduling \\
(SPDS) }
\label{sec:spds}

The simplest strategy for executing ensembles of workflows in the cloud is to use
provision resources statically and schedule workflow tasks dynamically. Given a 
budget $b$, deadline $d$, and the hourly price $p$ of a VM it is easy to calculate
the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before
the deadline:

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / d / p \rceil
\end{equation}

The SPDS algorithm statically provisions VMs at the start of the ensemble execution 
and keeps them running until the deadline is reached, or the budget runs out.
This provisioning plan has the advantage that it minimizes the number of provisioning 
and de-provisioning requests.

Once the VMs are provisioned, the tasks are mapped onto idle VMs using the dynamic
priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.

\begin{algorithm}
\caption{Priority-based scheduling algorithm for SPDS}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

Initially, the ready tasks from all workflows in the ensemble are added to a 
priority queue based on the priority of the workflow to which they belong. If
there are idle VMs available, and the priority queue is not empty, the next task
from the priority queue is submitted to a randomly chosen idle VM. The process is 
repeated until there are no idle VMs or the priority queue is empty. The scheduler
then waits for a task to finish, adds its ready children to the priority queue,
marks the VM as idle, and the entire process repeats until the deadline is reached.

This algorithm guarantees that tasks from lower priority workflows are
always deferred when higher-priority tasks are available, but lower-priority
tasks can still occupy idle VMs when higher-priority tasks are not available. 
However, because there is no preemption, long-running low-priority tasks may delay 
the execution of higher-priority tasks. In addition, tasks from low priority 
workflows may be executed even though there is no chance that those workflows 
will be completed within the current budget and deadline. Fig.~\ref{fig:spds-example} 
shows an example schedule generated using the SPDS algorithm. The figure illustrates 
how tasks from lower priority workflows backfill idle VMs when tasks from higher 
priority workflows are not available.

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example schedule generated using SPDS algorithm. Tasks are colored by workflow. }
\label{fig:spds-example}
\end{figure}

@@

The static provisioning approach of SPDS may not be optimal when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing
is that the number of resources can be changed in response to application demand. 
In policy-based provisioning (or auto-scaling) systems, the number of provisioned VMs 
is adjusted based on application demand as measured by metrics such as resource 
utilization and queue length. The DPDS algorithm uses resource utilization
to determine when to provision and de-provision VMs. The provisioner
module periodically computes utilization as the percentage of idle VMs. The
provisioner algorithm detects which VMs are finishing their hourly billing cycle
during the next provisioning period and decides whether they should continue
running, should be terminated or new VMs should be provisioned: see
Algorithm~\ref{alg:prov}.

Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate.

\begin{algorithm}
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold.
\Procedure{Provision}{}
    \State $V_C\gets set\ of\ VMs\ completing\ billing\ cycle$
    \State $n_T\gets\ 0$ \Comment{number of VMs to terminate} 
    \If{$ b - c < p * |V_C| \vee t > d $ }
    	\State $V_R\gets set\ of\ running\ VMs$
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$ 
    	\State $V_T\gets select\ n_T VMs\ to\ terminate\ from\ V_C$
    	\State \Call{Terminate}{$V_T$}
    \Else 
    	\If{$u>u_h$}
    		\State \Call{Start}{$new\ VM$}
    	\EndIf
    	\If{$u<u_l$}
    		\State $V_F\gets free\ (idle)\ VMs$
    		\State $n_T\gets \lceil|V_F|/2\rceil$ 
    		\State \Call{Terminate}{$V_T$}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The number of instances to terminate $n_T$ computed in line 6 is the number that
would overrun the budget if not terminated. On the other hand $n_T$ computed in
line 15 prevents terminating all the instances too quickly, but makes sure that
if there is only one instance left it should be terminated. However, the
termination process is not immediate: to make sure that VMs that have been
already paid for are not terminated prematurely the \textsc{Terminate} procedure
terminates only these instances that are close to the end of their billing
cycle. The remaining time threshold is calculated as a sum of provisioner
interval and estimated termination delay. This guarantees that no VMs can
overrun the assigned budget. To avoid uncontrolled increase of number of
instances, which may happen in the case of highly parallel workflows,
provisioner checks also the maximum scaling factor from the initial number of
provisioned VMs and prohibits starting new ones when this threshold is reached.

@@

DPDS algorithm does not use any information about workflow structure: it deals
only with scheduling of individual released/ready tasks using their
workflow priorities. If there are no high priority tasks in the ready queue,
then lower priority tasks are submitted to idle resources. This may lead to the
deficiency in the following scenario: if a long-running task of low
priority is submitted to the idle machine and then several high
priority tasks are released to the ready queue, they become delayed by the
execution of the low priority task. This usually does not cause a problem when
the number of tasks and resources is high, but when the execution approaches the
budget and deadline constraint, we have observed that tasks of low-priority
workflows may prevent higher priority workflows from completion. This
observation leads to the improved algorithm which we call Workflow-Aware DPDS
(WA-DPDS).
